Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-06-17 manuscript.Rmd
add a concrete example #34

>>>>

Both theorethical risk ($P(¬E|¬H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories.
To make these concepts less abstract, consider the following illustrative example of a theory and different experiments that test it.
Suppose the theory is about the cognitive phenomenon of "insight", and allow us to define it somewhat hand-waving as an cognitive abstraction that allows agents to consistently solve a well defined class of problems.
We pose the hypothesis that the following problem belongs to such class of insight problems:

> Use five matches (Ⅰ Ⅰ Ⅰ Ⅰ Ⅰ) to form the number eight.

We propose three experiments that differ in theoretical risk and detectability.
All take a sample of ten psychology students and present them the problem for brief span of time.
After that:

1. the experimenter explains that with roman-numerals the problem is easy to solve; if all students come up with the solution, she records it as evidence for the hypothesis.
2. the experimenter shows the solution "Ⅷ" and explains it; if all students come up with the solution, she records it as evidence for the hypothesis.
3. the experimenter does nothing; if all students come up with the solution, she records it as evidence for the hypothesis.

We argue that experiment 1 has high theoretical risk ($P(¬E_1|¬H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($¬H$), than presenting the insight that roman literals might be used, should not lead to all students solving the problem ($¬E_1$); the experiment has therefore high theoretical risk ($P(¬E_1|¬H)$)
Conversly, if insight is required to solve the problem ($H$), than it is probable to help all students to solve the problem ($E_1$); the experiment has therefore high detectability ($P(E_1|H)$).
The experiment 2 on the other hand, has low theoretical risk ($P(¬E_2|¬H)$).
Even if "insight" has nothing to do with solving the problem ($¬H$), there are other plausible reasons for observing the evidence ($E_2$), because the students can simply copy the solution, without any insight.
With regard to detectability experiment 1 and 2 differ in no obvious way.
Experiment 3, however, has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$); the experiment 3 has therefore low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms but high compared to the theoretical risk.
In the unlikely event that all 10 students lay the matches down in the form of the roman numeral Ⅷ ($E_3$) it is probably due to insight ($H$) and not by chance $P(¬E_2|¬H)$).
Of course, it comes to mind to allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.
We discuss the relation between statistical methods and theoretical risk in the section [Statistical Methods].
